Quantum Information

### 2024-11-21 examples/paper.tex
LOCC beginning, LOCC vs majorization

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage{amsthm,amsmath,amssymb}
\usepackage{tikz}
\usepackage[hidelinks,draft=false,colorlinks=true]{hyperref}
\usepackage[capitalize]{cleveref} % after hyperref!

@@

\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}

@@

\begin{theorem}[Ky-Fan]
  Let $\mathcal{H}$ be a Hilbert space, $k\in\mathbb{N}$, and 
  \begin{equation*}
    \mathcal{P}_k(\mathcal{H}):= \left\{ P\in\bounded{\mathcal{H}} \middle|  P = P^\dagger = P^2,\ \tr(P)=k\right\}.
  \end{equation*}
  Let $\rho\in \bounded{\mathcal{H}}$ Hermitian, and 
  \begin{equation*}
    \rho = \sum_{i=1}^{\dim\mathcal{H}} \lambda_i \ket{\Phi_i}\bra{\Phi_i}
  \end{equation*}
  be its eigen decomposition, with $\lambda_1 \geq \lambda_2 \geq \dots \geq \lambda_{\dim \mathcal{H}}$. Then 
  \begin{equation*}
    \sup_{P\in\mathcal{P}_k} \tr\{\rho P\} = \sum_{i=1}^k \lambda_i.
  \end{equation*}
\end{theorem}

@@

\begin{definition}[Majorization]
  Let $p,q\in\mathbb{R}^n$ be probability distributions. Let $p^\downarrow$ ($q^\downarrow$) be the vector obtained by listing the entries of $p$ ($q$) in descending order. We say that $p$ is majorized by $q$, and write $p\preccurlyeq q$, if for all $k=1,\dots, n$,
  \begin{equation*}
    \sum_{i=1}^k p^\downarrow_i \leq \sum_{i=1}^k q^\downarrow_i. 
  \end{equation*} 
\end{definition}
\end{document}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage{amsthm,amsmath,amssymb}
\usepackage{tikz}
\usepackage[hidelinks,draft=false,colorlinks=true]{hyperref}
\usepackage[capitalize]{cleveref} % after hyperref!
\usepackage{mathtools}  % for extendible arrows with text above

@@

\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\theoremstyle{definition}
\newtheorem{example}{Example}

@@

We want to understand which states are more entangled than others. Our strategy: identify a sufficiently large class of operations (this will be LOCC) that we declare not to increase entanglement, then we say that $\ket{\Psi}$ is more entangled than $\ket{\Phi}$ iff we can transform $\ket{\Psi}$ to $\ket{\Phi}$ via those operations.

Which operations shall we choose? Quantum operations of (pure) states, thus they will consist of unitaries and measurements. Minimal requirement: do not transform separable states to entangled states. Unitaries are easy. Only product unitaries are allowed, as
\begin{proposition}
  Let $U\in \bounded{\mathcal{H}\otimes\mathcal{K}}$. Then $U (\ket{\psi} \otimes \ket{\phi})$ is a product state for all $\ket{\psi}\in\mathcal{H}$, $\ket{\phi}\in\mathcal{K}$, if and only if $U$ is a product, i.e., there are $V\in\bounded{\mathcal{H}}$ and $ W\in \bounded{\mathcal{K}}$ unitaries such that $U = V\otimes W$ .
\end{proposition}
\begin{proof} Exercise.
  %If $U = V\otimes W$, then $U (\ket{\psi} \otimes \ket{\phi}) = V\ket{\psi} \otimes W\ket{\phi}$, it is a product state. If $U (\ket{\psi} \otimes \ket{\phi}) = \ket{\psi'} \otimes \ket{\phi'}$, for all $\ket{\psi}$ and $\ket{\phi}$, then due to linearity of the l.h.s.,  
\end{proof}

\begin{theorem}[Ky-Fan]\label{thm:Ky-Fan}
  Let $\mathcal{H}$ be a Hilbert space, $k\in\mathbb{N}$, and 
  \begin{equation*}
    \mathcal{P}_k(\mathcal{H}):= \left\{ P\in\bounded{\mathcal{H}} \middle|  P = P^\dagger = P^2,\ \tr(P)=k\right\}.
  \end{equation*}
  Let $\rho\in \bounded{\mathcal{H}}$ Hermitian, and 
  \begin{equation*}
    \rho = \sum_{i=1}^{\dim\mathcal{H}} \lambda_i \ket{\Phi_i}\bra{\Phi_i}
  \end{equation*}
  be its eigen decomposition, with $\lambda_1 \geq \lambda_2 \geq \dots \geq \lambda_{\dim \mathcal{H}}$. Then 
  \begin{equation*}
    \sup_{P\in\mathcal{P}_k} \tr\{\rho P\} = \sum_{i=1}^k \lambda_i.
  \end{equation*}
\end{theorem}

@@

\begin{definition}[Majorization]
  Let $p,q\in\mathbb{R}^n$ be probability distributions. Let $p^\downarrow$ ($q^\downarrow$) be the vector obtained by listing the entries of $p$ ($q$) in descending order. We say that $p$ is majorized by $q$, and write $p\preccurlyeq q$, if for all $k=1,\dots, n$,
  \begin{equation*}
    \sum_{i=1}^k p^\downarrow_i \leq \sum_{i=1}^k q^\downarrow_i. 
  \end{equation*} 
\end{definition}

{\color{red} TODO: Here it is better to define Schmidt coefficient with zero in it.}

\begin{proposition}
  Let $\ket{\Psi},\ket{\Phi}\in \mathcal{H}\otimes \mathcal{K}$. Let the squared Schmidt coefficients of $\ket{\Psi}$ be $p$, and the squared Schmidt coefficients of $\ket{\Phi}$ be $q$.  If $\ket{\Psi} \xrightarrow{LOCC} \ket{\Psi}$, then $q\preccurlyeq p$.
\end{proposition}

\begin{proof}
  Let $\rho = \tr_{\mathcal{H}}\ket{\Psi}\bra{\Psi}$ and $\eta = \tr_{\mathcal{H}}\ket{\Psi}\bra{\Psi}$. Then the eigenvalues of $\rho$ are precisely the entries of $p$ and the eigenvalues of $\eta$ are the entries of $q$.  Using {\color{red} MISSING},  $\ket{\Psi} \xrightarrow{LOCC} \ket{\Psi}$ if and only if
  \begin{equation*}
    \sqrt{r_i}\ket{\Phi} = M_i \otimes U_i \ket{\Psi},
  \end{equation*}
  for a measurement $\{M_i\}_{i=1}^n$, a probability distribution $r\in\mathbb{R}^n$ and some unitaries $\{U_i\}_{i=1}^n$. Therefore
  \begin{equation*}
    r_i \eta = U_i \rho_i U_i^\dagger,
  \end{equation*} 
  where $\rho_i = \tr_{\mathcal{H}} \{ (M_i^\dagger M_i \otimes \id)\ket{\Psi}\bra{\Psi}\}$,  and thus 
  \begin{equation*}
    \sup_{P\in\mathcal{P}_k} \tr\{P \eta\} = \sup_{P\in\mathcal{P}_k} \sum_i  \tr\{r_i P \eta\} = \sup_{P\in\mathcal{P}_k} \sum_i \tr\{PU_i \rho_i U_i^\dagger \} \leq \sum_i \sup_{P\in\mathcal{P}_k} \tr\{U_i^\dagger P U_i \rho\} = \sum_i \sup_{P\in\mathcal{P}_k} \tr\{P \rho_i\} = \sup_{P\in\mathcal{P}_k} \tr\{P\rho\}.  
  \end{equation*}
  Using now \cref{thm:Ky-Fan}, we obtain that $p\preccurlyeq q$.
\end{proof}

Let us try to understand now a \emph{sufficient} criterion for $\ket{\Psi}\xrightarrow{LOCC}\ket{\Phi}$.

\begin{example}
  Let $\ket{\Psi} =  \sqrt{1/2} \ket{00} + \sqrt{1/2}\ket{11}$ and $\ket{\Phi} = \sqrt{2/3} \ket{00} + \sqrt{1/3}\ket{11}$. Notice that 
  \begin{equation*}
    M_0 = \left( \begin{matrix}
      \sqrt{2/3} & 0 \\ 0 & \sqrt{1/3}
    \end{matrix} \right),
  \end{equation*}
  is a matrix such that $(M_0\otimes \id) \ket{\Psi} = \ket{\Phi}$. Moreover, we can complete it to a measurement, for example,  with the matrix
  \begin{equation*}
    M_1 = \left( \begin{matrix}
      \sqrt{1/3} & 0 \\ 0 & \sqrt{2/3}
    \end{matrix} \right),
  \end{equation*}
  i.e., $M_0^\dagger M_0 + M_1^\dagger M_1  = \id$.  Notice that $(M_1\otimes \id) \ket{\Psi} = X\otimes X\ket{\Phi}$. Therefore the measurement $N_i = X^i M_i$ ($i=1,2$) is such that 
  \begin{equation*}
    M_i \otimes X^i \ket{\Psi} = \sqrt{p_i} \ket{\Phi},
  \end{equation*}
  for some probability distribution $p_i$, i.e., $\ket{\Psi}\xrightarrow{LOCC}\ket{\Phi}$.
\end{example}
The essence of this example is that the squared Schmidt coefficients of the states are related to each other as
\begin{equation*}
   \left(\begin{matrix} 1/2 \\ 1/2 \end{matrix}\right) = \frac{1}{2} \cdot (\id + X) \cdot \left(\begin{matrix} 1/3 \\ 2/3 \end{matrix}\right).
\end{equation*}
This construction can be easily generalized for the case when the squared Schmidt coefficients of $\ket{\Psi}$ can be obtained by mixing permutations of the squared Schmidt coefficients of $\ket{\Phi}$. Before we show how to obtain the LOCC operations for this generalized case, notice that this constraint is equivalent with majorization:

\begin{lemma}\label{lem:majorization=>mixing_permutation}
  Let $p,q\in\mathbb{R}^n$ be probability distributions. If $p\preccurlyeq q$, then there is a probability distribution $r\in\mathbb{R}^{n!}$ such that 
  \begin{equation*}
    p  = \sum_{\pi \in S_n} r_{\pi} \pi q,
  \end{equation*}
  where $S_n\subseteq \bounded{\mathbb{R}^n}$ is the set of $n\times n$ permutation matrices. 
\end{lemma}

\begin{proof}
  Exercise.
\end{proof}

{\color{red} Again, number of non-zero Schmidt coeffs might be different!}
\begin{proposition}
  Let $\ket{\Psi},\ket{\Phi}\in \mathcal{H}\otimes \mathcal{K}$. Let the squared Schmidt coefficients of $\ket{\Psi}$ be $p$, and the squared Schmidt coefficients of $\ket{\Phi}$ be $q$. If $p\preccurlyeq q$, then $\ket{\Psi}\xrightarrow{LOCC}\ket{\Phi}$. 
\end{proposition}

\begin{proof}
  Using \cref{lem:majorization=>mixing_permutation}, we obtain that there is a probability distribution $r$ such that $p  = \sum_{\pi \in S_n} r_{\pi} \pi q$, where $S_n$ is the group of $n\times n$ permutation matrices. Let the Schmidt decomposition of the states be 
  \begin{align*}
    \ket{\Psi} = \sum_{i=1}^n p_i \cdot \ket{l_i} \otimes \ket{r_i}\\
    \ket{\Phi} = \sum_{i=1}^n q_i \cdot \ket{\hat{l}_i} \otimes \ket{\hat{r}_i},    
  \end{align*}
  and let us define operators $M_{\pi}$ as 
  \begin{equation*}
    M_{\pi} = \sum_i \left( r_\pi \frac{q_{\pi(i)}}{p_i}\right)^{1/2} \ket{\hat{l}_{\pi(i)}} \bra{l_i}. 
  \end{equation*}
  These operators form a POVM as
  \begin{equation*}
    \sum_{\pi} M_\pi^\dagger M_\pi = \sum_{ij}\sum_\pi  r_\pi \frac{q_{\pi(i)}}{p_i} \ket{l_i}\scalprod{\hat{l}_{\pi(i)}}{\hat{l}_{\pi(j)}}\bra{l_j} = \sum_{i}\sum_\pi  r_\pi \frac{q_{\pi(i)}}{p_i} \ket{l_i}\bra{l_i}  = \sum_{i} \ket{l_i}\bra{l_i}  = \id.
  \end{equation*}
  Note as well that 
  \begin{equation*}
    (M_\pi \otimes \id) \ket{\Psi} = \sum_i  \sqrt{r_\pi q_{\pi(i)}} \ket{\hat{l}_{\pi(i)}}\otimes \ket{r_i}, 
  \end{equation*}
  and thus, setting $V_{\pi} = \sum_i \ket{r_{\pi(i)}}\bra{r_i}$, $V_\pi$ is a unitary and 
  \begin{equation*}
    (M_\pi \otimes V_\pi) \ket{\Psi} = \sum_i  \sqrt{r_\pi q_{\pi(i)}} \ket{\hat{l}_{\pi(i)}}\otimes \ket{r_{\pi(i)}} = \sqrt{r_\pi} \ket{\Phi}, 
  \end{equation*}
  i.e., we have shown that $\ket{\Psi}\xrightarrow{LOCC}\ket{\Phi}$.
\end{proof}

We have thus seen that
\begin{theorem}
  Let $\ket{\Psi},\ket{\Phi}\in \mathcal{H}\otimes \mathcal{K}$. Let the squared Schmidt coefficients of $\ket{\Psi}$ be $p$, and the squared Schmidt coefficients of $\ket{\Phi}$ be $q$. Then $\ket{\Psi}\xrightarrow{LOCC}\ket{\Phi}$ iff  $p\preccurlyeq q$.   
\end{theorem}

\end{document}
